---
title: "Correlations between multiply censored variables arising from adjustment"
author: "Herb Susmann"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{bayesplot}
---

```{r, include = FALSE}
library(censcor)
set.seed(1)
```

Often in environmental health datasets censored variables are adjusted by another non-censored variable. For example, measurements of two chemicals in a sample ($x$ and $y$) might be normalized by dividing by the volume of each sample ($z$). If $x$ and $y$ were singly censored, dividing them by $z$ leads to variables $x_{adj}$ and $y_{adj}$ that are censored at multiple points.

[Newton and Rudel (2007)](https://www.ncbi.nlm.nih.gov/pubmed/17265951) describe a method for estimating the Pearson correlation between two censored variables that are adjusted by a third, non-censored variable. The approach incorporates information about the correlations between $x$, $y$, and $z$, and is based on the identity

$$
\mathrm{cor}(x-z,y-z) = \frac{\sigma_{xy} - \sigma_{xz} - \sigma_{yz} + \sigma_z^2}{\big\{(\sigma_x^2 + \sigma_z^2 -2 \sigma_{xz})(\sigma_y^2 + \sigma_z^2 -2\sigma_{yz})\big\}^{0.5}}
$$

You can use this method in `censcor` by using the `adj` argument to input a formula describing the adjustment column. Note that we still the _unadjusted_ values for $x$ and $y$.

```{r, results = 'hide'}
d1 <- generate_censored_data(N = 100,        # Number of data points
                             rho = 0.5,      # true correlation between x and y
                             L_prob = 0.2,   # probability of censoring
                             direction = -1, # direction of censoring (left)
                             adj = TRUE,     # generate an adjustment variable
                             adj_sigma = 1)  # standard deviation of the adjustment variable

fit <- censcor(x | cens(x_cens) ~ y | cens(y_cens), d1, adj = ~adj, iter = 600)

```

The estimated correlation between $x$ and $y$ as adjusted by $z$ (that is, the correlation between $x-z$ and $y-z$) is reported as `rho_adj`. The theoretical value is

$$
\rho_{adj} = \frac{\rho_{xy} + \sigma_z}{1 + \sigma_z}
$$

In this example, we set $\rho_{xy}=0.5$ and $\sigma_z=1$, so we expect $\rho_{adj}$ to be $0.75$.

```{r}
fit
```

```{r}
library(bayesplot)

mcmc_areas(as.matrix(fit), pars = "rho_adj")

```