---
title: "Introduction to censcor"
author: "Herb Susmann"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(censcor)
```

The `censcor` package provides routines for performing Bayesian estimation of the correlation between two censored variables. The inference is conducted with `Stan`, which uses the `No-U-Turn Hamiltonian Monte-Carlo` algorithm for drawing from the posterior distribution.

## Usage

The main function is `censcor`, which computes correlations between two variables. `censcor` takes as arguments a formula describing the correlation to estimate and a data frame. The formula syntax is modeled after the censoring syntax in `brms`.

The formula syntax is:

```{r, eval = FALSE}
censcor(x | cens(x_cens, x_upper_bound) ~ y | cens(y_cens, y_upper_bound), df)
```

Let's look at some examples of how to specify different types of censoring. The `generate_censored_data` function generates a data frame of random values for testing the package.

### No censoring

The most basic estimation you can perform is between two variables that are not censored. In this example, the correlation between the `x` and `y` columns of `df` is estimated.

```{r, eval=FALSE}
# Generate 100 x and y values with a 0.5 correlation coefficient and no censoring
df <- generate_censored_data(N = 100, rho = 0.5, L_prob = 0)

fit <- censcor(x ~ y, df)
```

### Left censoring
Censoring is described with the `cens` formula function. A separate column is needed for each censored variable to indicate which values are censored, with -1 indicating left censored, 1 indicating right censored, and 2 indicating interval censored. Censored values in `x` and `y` should be replaced with the censoring limit.

In this example, both `x` and `y` are left-censored. `x_cens` and `y_cens` indicate which rows are censored by having the value -1.

```{r, eval=FALSE}
# df example:
# x  x_cens   y   y_cens
# 2  -1       1   0        # x is left-censored below 2, y is not censored
# 0   0       4  -1        # x is not censored, y is left-censored below 2
# ...

df <- generate_censored_data(N = 100, rho = 0.5, L_prob = 0.5, direction = -1)

fit <- censcor(x | cens(x_cens) ~ y | cens(y_cens), df)
```

### Right censoring

The syntax for right-censoring is the same as before. `x_cens` and `y_cens` contain 1s to indicate right-censored values.

```{r, eval=FALSE}
# df example:
# x  x_cens  y   y_cens
# 5  1       1   0        # x is right-censored above 5, y is not censored
# 0  0       2   1        # x is not censored, y is right-censored above 2
# ...

df <- generate_censored_data(N = 100, rho = 0.5, L_prob = 0.5, direction = 1)

fit <- censcor(x | cens(x_cens) ~ y | cens(y_cens), df)
```

### Interval censoring

Interval censored values have lower and upper bounds. In this case, `x` and `y` contain the lower interval limit, and separate columns are used to hold the uper interval limit and are specified as the second argument to `cens`.


```{r, eval=FALSE}
# df example:
# x   x_upper  x_cens   y   y_upper   y_cens
# 1   5        2        1   1         0        # x is interval censored between 1 and 5; y is not censored
# ...

df <- generate_censored_data(N = 100, rho = 0.5, L_prob = 0.5, direction = 2)

fit <- censcor(x | cens(x_cens, x_upper) ~ y | cens(y_cens, y_upper), df)
```


### Mixed censoring

Left, right, and interval censoring can be combined within a data frame.

```{r}
# df example:
#  x   x_upper  x_cens   y   y_upper   y_cens
#  1   5        2        1   1         0        # x is interval censored between 1 and 5; y is not censored
#  3   3       -1        5   1         1        # x left censored below 3, y is right censored above 5
# -1  -1        1        1   5         2        # x is right censored above -1, y is interval censored between 1 and 5
# ...

df <- generate_censored_data(N = 100, rho = 0.5, L_prob = 0.5, direction = c(-1, 1, 2))

fit <- censcor(x | cens(x_cens, x_upper) ~ y | cens(y_cens, y_upper), df)

```

### Output

`censcor` returns an `rstan::stanfit` object for several fitted parameters.

- `mu_x` is the estimated mean of the `x` variable
- `sigma_x` is the estimated standard deviation of the `x` variable
- `mu_y` is the estimated mean of the `y` variable
- `sigma_y` is the estimated standard deviation of the `y` variable
- `rho` is the estimated correlation between `x` and `y`

```{r}
library(rstan)

df <- generate_censored_data(N = 100, rho = 0.5, L_prob = 0.5, direction = -1)

fit <- censcor(x | cens(x_cens, x_upper) ~ y | cens(y_cens, y_upper), df)

summary(fit)

```

You can visualize the estimates using the `bayesplot` package:

```{r}
library(bayesplot)

mcmc_dens(fit, pars = c("rho"))
```
